Analysis of US Campaign donations by Pierre Hentges
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

library(ggplot2)
library(dplyr)
library(reshape2)
library(gridExtra)
#install.packages("treemapify")
library(treemapify)
library(tidyverse)
library(ggfittext)
#install.packages("ggalluvial")
library(ggalluvial)
#install.packages("ggrepel")
library(ggrepel)

```

```{r echo=FALSE, misc_variables}
#define some variables that will bee used in data cleaning and analysis
#Define candidates of each party
Democrats <- c("Clinton, Hillary Rodham", "Lessig, Lawrence", 
               "O'Malley, Martin Joseph", "Sanders, Bernard",
               "Webb, James Henry Jr." )
Republicans <- c("Trump, Donald J.", "Bush, Jeb", "Carson, Benjamin S.", 
                 "Christie, Christopher J.","Cruz, Rafael Edward 'Ted'", 
                 "Fiorina, Carly", "Gilmore, James S III", "Graham, Lindsey O.", 
                 "Huckabee, Mike", "Jindal, Bobby","Kasich, John R.", 
                 "Pataki, George E.", "Paul, Rand", "Perry, James R. (Rick)", 
                 "Rubio, Marco", "Santorum, Richard J.", "Walker, Scott" )
Libertarian <- c("Johnson, Gary") 
Green <- c("Stein, Jill")
Independent <- c("McMullin, Evan") 

#US states
states <- c("AL","AK","AZ","AR","CA","CO","CT","DE","FL","GA","HI","ID","IL",
            "IN","IA","KS","KY","LA","ME","MD","MA","MI","MN","MS","MO","MT",
            "NE","NV","NH","NJ","NM","NY","NC","ND","OH","OK","OR","PA","RI",
            "SC","SD","TN","TX","UT","VT","VA","WA","WV","WI","WY")

#Swing states  Colorado, Florida, Iowa, Michigan, Minnesota, Ohio, Nevada, New 
#Hampshire, North Carolina, Pennsylvania, Virginia, and Wisconsin 
#according to https://en.wikipedia.org/wiki/Swing_state
SwingStates <- c("CO","FL","IA","MI","MN", "OH", "NV", "NH", "NC", "PA", 
                 "VA", "WI")

```

> The dataset to be analysed is of campaign contributions to the US 
presidential elections from 2016. This includes donations in both the primary 
and general election campaigns, to candidates of all parties and across all US 
states. 

> The dataset was downloaded from http://fec.gov and cleaned at the start, then 
saved as two .csv files (campaign2016data.csv, returned2016data.csv). Given the 
size of this data set -over 7 million records- most of the variables that will 
not be analysed have been removed. The cleaned data are submitted in the zip 
archive and are loaded in an early chunk of R code, while the two R chunks that
clean the original file have been set to not run by default ("include=FALSE").


```{r eval=FALSE, include=FALSE}
#define functions used in data cleaning

#function to derive candidate's party affiliation
partyAffFunction <- function(CandtName){ #argument: variable with Candidate name
CandtParty <- as.factor(ifelse(CandtName %in% Democrats,"Democrats", 
                               ifelse(CandtName %in% Republicans,"Republicans",
              ifelse(CandtName %in% Libertarian,"Libertarian",
                     ifelse(CandtName %in% Green,"Green",
                            ifelse(CandtName %in% Independent,"Independent",
                                   "NA"))))))
return(CandtParty)
}


#function passing list to levels() to clean up freeform occupation data, 
#adapted from #https://stackoverflow.com/a/47210732
CleanOccupations <- function(Occupations){ #arg column w/ factorised occupations 
levels(Occupations)   <- list( 
  "Retired"=c('RETIRED','RETIREE'), 
  "Unemployed"=c("NOT EMPLOYED","NOT IN WORKFORCE","UNEMPLOYMENT","UNEMPOLYED",
                 "UNEMPLOYED","UNEMPLYE","UNEMPLYD","NOT EMPLOYED",
                 "NOT-EMPLOYED","N/A","NONE"),
  "Info Witheld"=c("INFORMATION REQUESTED","INFORMATION REQUESTED PER BEST 
                EFFORTS",""," ","REFUSED","CONDUIT TOTAL LISTED IN AGG. FIELD"),
  "Lawyer"=c("LAWYER","ATTORNEY","ATTORMEY","LAYWER","ATTONEY","DEFENDER"),
  "Teacher"=c("TEACHING","EDUCATION","TEACHER","SCHOOL","EDUCATOR"),
  "Academic"=c("UNIVERSITY","UNIV","COLLEGE","PHD","PROF","COLLEGE PROFESSOR",
          "PROFESOR", "PROFESSOR","LECTURER", "ACADEMIC", "LECTURER","FACULTY", 
      "SCIENTIST","SCHOLAR","LINGUIST","PHILOSOPHER","ARCHAEOLOGIST","CHEMIST",
      "HISTORIAN","RESEARCH","BIOLOGIST","PHYSICIST","CHEMIST","RESEARCHER"),
  "Nurse"=c("NURSE","R.N.","RN","REGISTERED NURSE","NURSE PRACTITIONER"),
  "Doctor"=c("ORTHOPEDIST","PEDIATRICS","GASTROENTEROLOGY","PULMONOLOGIST",
             "OPHTHAMOLOGIST","OPHTHALMOLOGY","CLINICIAN","PSYCHIATRY",
            "ANESTHESIOLOGY","NEONATOLOGIST","ANESTHETIST","GASTROENTEROLOGIST",
            "OPTHALMOLOGIST","ENDODONTIST","PATHOLOGIST","PATHOLOGIST",
            "PERIODONTIST","PHYSCIAN","ONCOLOGIST","OPTHAMOLOGIST","DENTIST",
            "ANESTHETIST","OPHTHALMOLOGIST","PHYSICAN","DOCTOR","DENTIST",
            "PHYSICIAN","SURGEON","ANESTHESIOLOGIST","MD","ORTHODONTIST",
            "OPTOMETRIST","PEDIATRICIAN","CARDIOLOGIST","PSYCHIATRIST",
            "GASTROENTEROLOGY","RADIOLOGY","UROLOGIST","DERMATOLOGIST",
            "PODIATRIST","VETERINARIAN","M.D."),
  'Self-employed'=c("SELF-EMPLOYED","SELF EMPLOYED","SELF"),
  "Homemaker"=c("HOMEMAKER","HOME MAKER","HOUSE WIFE","WIFE","WIDOW","AT HOME",
                "MOM","MOTHER","HOUSEWIFE"),
"Finance"=c("MORTGAGE","BANK","BANKER","BANKING", "LENDER","CPA", "C. P. A.", 
            "INVESTING", "FINANCIER", "INVESTER", "INVESTOR","INVESTMENTS",
            "INVESTMENT", "FINANCE", "ANALYST", "ACCOUNTANT", "BOOKKEEPER",
            "UNDERWRITING","CPA","FINANCIAL ADVISOR","FINANCIAL","INSURANCE"),
"Real Estate"=c("REAL ESTATE", "REALTY", "REAL-ESTATE", "REALESTATE", "REAL 
                ESTATE", "REAL ESTATE OWNER", "DEVELOPMENT", "DEVELOPER","REAL 
                ESTATE BROKER","REALTOR"),
"Senior manager" = c("CEO","CHIEF OPERATING OFFICER","C. E. O.","C.E.O",
                     "C.F.O.","C.F.O","CFO","C.I.O.", "CPO","CAO","CIO", "CMO",
                     "CTO", "C.O.O.", "COB","COO","CCO","CFP","GM","DIRECTOR",
                    "CHIEF EXECUTIVE OFFICER", "EXECUTIVE","EXECUTIVE DIRECTOR", 
                     "EXEC","MANAGEMENT", "MANAGER","MANGER","MANAGING",
                    "PRESIDENT","VICE PRESIDENT","VICE-CHAIR","CHAIRMAN",
                    "PROJECT MANAGER","CHARMAN","CHARIMAN","VC","VP", "BOARD",
                    "VICE", "BOSS","MGT", "MGR","TRUSTEE", "BOARD MEMBER",
                    "TRUSTEE", "CHAIR","VICE CHAIR","DIRECTOR","PRESIDENT","VP",
                    "CHAIR","V.P.","PROGRAM MANAGER"), 
"Programmer&IT" = c("PROGRAMMER","COMPUTER PROGRAMMER","IT","SOFTWARE 
                    DEVELOPER","SOFTWARE ENGINEER","WEB DEVELOPER"),
"Farmer" = c("FARMER","RANCHER"),
"Admin" = c("ADMINISTRATIVE ASSISTANT","OFFICE MANAGER","PARALEGAL",
            "SECRETARY","ADMINISTRATOR"),
"Arts" = c("ACTOR","ARTIST","MUSICIAN","PHOTOGRAPHER","WRITER"),
"Media" = c("DESIGNER", "EDITOR","GRAPHIC DESIGNER"),
"Student" = c("STUDENT","GRADUATE STUDENT"),
"Business owner" = c("BUSINESS OWNER","OWNER","SMALL BUSINESS OWNER",
                     "BUSINESSMAN","BUSINESSWOMAN","CAPITALIST", "SHAREHOLDER", 
                     "VENTURE","ENTREPRENEUR", "ENTREPENEUR", "WHOLESALE",
                     "EXPORTER","SELLER","WHOLESALER","IMPORTER","DISTRIBUTOR",
                     "MERCHANT"),
"Sales & marketing" = c("SALES","MARKETING","MRKTNG","RETAIL","E-COMMERCE",
                        "ECOMMERCE", "DISTRIBUTION", "ADVERTISING"),
"Driver"="DRIVER",
"Electrician"="ELECTRICIAN",
"Engineer"="ENGINEER",
"Truck driver"="TRUCK DRIVER",
"Pilot"="PILOT",
"Mechanic"="MECHANIC",
"Librarian"="LIBRARIAN",
"Psychologist"=c("PSYCHOLOGIST", "PSYCHOTHERAPIST"),
"Architect" = "ARCHITECT",
"Disabled" = c("DISABLED","CONDUIT TOTAL LISTED IN AGG. FIELD DISABLED"),
"Construction"="CONSTRUCTION",
"Consultant"="CONSULTANT",
"Contractor"="CONTRACTOR",
"Pastor"="PASTOR",
"Pharmacist"="PHARMACIST",
"Social Worker"="SOCIAL WORKER")
return(Occupations)
}
#this function was heavily modified from 
#https://github.com/datasciencedojo/DataMiningFEC/blob/master/6%20Bucketing%20Occupation%20Groups.R
```



```{r eval=FALSE, include=FALSE}

#downloaded archive file containing a 1.4 GiB .csv file with data on 
#contributions to the 2016 US presidential election campaigns from 
#http://fec.gov/disclosurep/pnational.do&sa=D&ust=1548605508118000
# 2016_P00000001-ALL.zip, which contains 2016_P00000001-ALL.csv
#The column format is specified in the file CONTRIBUTOR_FORMAT.txt

#set working directory
#setwd() 


#Column names to be specified by passing the col.names parameter to read.csv()
column_names=c("cmte_id","cand_id","Candidate","Contributor_Name","City",
               "State","ZIP","contbr_employer","contbr_occupation",
               "Contribution","Date","receipt_desc","memo_cd","memo_text",
               "form_tp","file_num","tran_id","Election_Type","")

#To efficiently load the .csv file, I will drop (using colClasses = "NULL") and 
#specify the desired column classes ("factor", "numeric","character")

column_classes=c("NULL","NULL","factor","factor","factor",
                 "factor","NULL", "NULL","factor",
                 "numeric","character","factor","factor","factor",
                 "NULL","NULL","NULL","factor","NULL")

#there is a trailing comma at the end of rows not present in the header (column 
#names). to avoid a mismatch in the structure, ignoring the header and first 
#line, and specifying "NULL" for the final column
campaign2016 <- read.csv(2016_P00000001-ALL.csv',
  header = F,
  col.names = column_names, 
  colClasses = column_classes,skip = 1)

#formatting the contribution dates correctly
campaign2016$Date <- as.Date(campaign2016$Date,format="%d-%b-%Y")

#add new column with Party affiliation, based on candidate name
campaign2016$Party <- partyAffFunction(campaign2016$Candidate)

#Add column with candidate surname only (to avoid lenghty lables), 
#inspired by https://stackoverflow.com/a/31925893
campaign2016$CandidateSurname <- as.factor(sub("(\\w+).*", "\\1", 
                                               campaign2016$Candidate))

#Rename Factor levels of general and primaries, NA for other types
campaign2016$Election_Type <- recode(campaign2016$Election_Type, 
                                     G2016 = "General", P2016 = "Primary",  
                                     .default = NA_character_)

#only keep general and primaries
campaign2016 <- subset(campaign2016,!is.na(Election_Type))

#only keep contributions from one of the 50 US states
campaign2016 <- subset(campaign2016,State %in% states)
campaign2016$State <- factor(campaign2016$State) #removing unused factor levels

#clean occupation using function
campaign2016$Occupation <- CleanOccupations(campaign2016$contbr_occupation)


#separate dataframe for returned (negative) contributions, then drop unneeded 
#columns from main dataframe
returned2016 <- subset(campaign2016, Contribution<0)

#drop unneeded columns to reduce memory usage
campaign2016 <- subset(campaign2016, select = -c(contbr_occupation,receipt_desc, 
                                                 memo_cd, memo_text))

write.csv(campaign2016, file = "campaign2016data.csv",row.names=FALSE)
write.csv(returned2016, file = "returned2016data.csv",row.names=FALSE)


```



```{r echo=FALSE, Load_the_Data}
#set working directory to location of campaign2016data.csv 
#and returned2016data.csv

#setwd('')

#reading in cleaned csv files
column_names=c("Candidate","Contributor_Name","City","State","Contribution",
               "Date","Election_Type","Party","CandidateSurname","Occupation")
column_classes=c("factor","character","factor","factor","numeric",
                 "Date","factor","factor","factor","factor")
campaign2016 <- read.csv('campaign2016data.csv',
                         header = F,
                         col.names = column_names, 
                         colClasses = column_classes, skip = 1)

column_names=c("Candidate","Contributor_Name","City","State",
               "contbr_occupation","Contribution",
               "Date","receipt_desc","memo_cd","memo_text","Election_Type",
               "Party","CandidateSurname","Occupation")
column_classes=c("factor", "character", "factor","factor",
                 "factor","numeric",
                 "Date","character","character","character","factor",
                 "factor","factor","factor")
returned2016 <- read.csv('returned2016data.csv',
                         header = F,
                         col.names = column_names, 
                         colClasses = column_classes,skip = 1)
```


# Univariate Plots Section


```{r echo=FALSE, Univariate_Plots_dim}
dim(campaign2016)
```

The cleaned data set consists of 7.3 million contributions, with 10 variables. 
Let's check them out more closely.

```{r echo=FALSE, Univariate_Plots_str}
str(campaign2016)
```


```{r echo=FALSE, Univariate_Plots_summary}
summary(campaign2016)
```

* many of the variables are categorical (or factorised), including the 25 
presidential candidates and their party, whether the contribution was made in 
the primaries or general election, information on the contributor (state, 
occupation)
* the only numeric variable is Contribution amount, and the date of the 
contribution. These will be studied first.

* the majority of contributions (4.7 million) were in the primaries, 2.6 million 
for general election
* Clinton and Sanders got a lot more contributions than Trump, also the 
Democrats got far more contributions than the Republicans
* a number of variables (contributor name,address,employer,receipt comments etc)
have been dropped as they won't be analysed.


I'll start by looking more closely at the Contributions variable. The 
Contributions variable has a surprising range: from negative (-\$93308) to 
\$4.904.861. 

I'll disregard the negative contributions for now and look at the 
distribtion of positive contributions. The negative contributions have been 
split off in a separate dataframe, to be analysed further down. This dataframe 
which has additional columns that have been dropped from the main dataframe for 
memory efficiency reasons.


```{r echo=FALSE, Univariate_Plots_positive_contribution_summary}
#removing negative and 0 contributions - there is a seperate dataframe for them
campaign2016 <- subset(campaign2016,Contribution>0) 
summary(campaign2016$Contribution)
```


*The interquartile range (IQR) of positive contributions is $72, and varies 
between \$15 to \$87 - but range goes up to a maximum of $4.904.861 - it looks 
like there could be some outliers. Asymmetric shape is suggested by the 
mean (\$124) being much higher than the Median (\$28).
histograms to look at the shape:

```{r echo=FALSE, warning=FALSE, Univariate_Plots_histograns_log}
p1 <- qplot(data = campaign2016, x = Contribution, bins = 100, ylab = "count")
p2 <- p1 + scale_x_log10(breaks = c(0.01,0.1,1,10,100,1000,10000,1000000)) +
                         labs(x="Contribution (log scale)")
p3 <- p2 +  scale_y_log10() + labs(y = "Count (log scale)")
grid.arrange(p1,p2,p3)
```

The campaign contribution data are extremely right-skewed: in a histogram with
100 bins, 7 million donations are grouped in a single bar.
Using a logarithmic x scale: much of the data are now visible over 2 orders of 
magnitude, but extreme values are still not discernible.
Using logarithmic scales for both the x and y axes makes outliers visible. The 
shape now approximates a symmetric distribution more. The donations of several 
million \$ are clearly outliers as they are 2 orders of magnitude away from the
rest of the data. One could argue that the small number of donations below $0.10
could also be classified as outliers since they are separated from the rest of 
the data.

To better understand the values at the very top and bottom percentiles I will 
look more closely at these outliers.


```{r echo=FALSE, Univariate_Plots_quantiles}
quantile(campaign2016$Contribution, probs = c(0.0001, 0.001, 0.95, 0.99, 0.999,
                                              0.9999,0.99999, 0.999999, 1))

```

It is noticeable how the top quantiles of 99.9% and 99.99%  well below $10K. 
Looking at quantiles, 0.1% to 99.9% corresponds to \$1 - \$2700 ; and 0.01% to 
99.99% corresponds to \$0.80 - \$5400. These may be related to campaign finance 
legal limitations.

```{r echo=FALSE, warning=FALSE, Univariate_Plots_left_quantiles}
# looking at 0.1% and 0.01% quantiles using strict (< $1) 
# and non-strict inequalities

p1 <- qplot(data = subset(campaign2016, Contribution <= 1), x = Contribution, 
            xlab = "Contributions ≤ $1", ylab = "count", bins = 100)
p2 <- qplot(data = subset(campaign2016, Contribution < 1), x = Contribution, 
            xlab = "Contributions < $1", ylab = "count", bins = 100)
grid.arrange(p1,p2)
```

There are over 5,000 contributions of \$1 but only few donations below $1. Their 
usefulness for this study is questionable. As these take up 2 units on a 
log-scale x-axis if the full range is plotted, I will consider removing them 
from the analysis.

```{r echo=FALSE, warning=FALSE, Univariate_Plots_right_quantiles}
#looking at 99.9% and 99.99% quantiles using strict ( > $2700) and 
#non-strict inequalities

p1 <- qplot(data = subset(campaign2016, Contribution >= 2700), 
            x = Contribution, 
            xlab="Contributions ≥ $2700 (log scale)", 
            ylab = "count", 
            bins = 100) +
  scale_x_log10(breaks = c(2700, 5400, 12000, 100000, 1000000))
p2 <- qplot(data = subset(campaign2016, Contribution > 2700),
            x = Contribution,
            xlab = "Contributions > $2700 (log scale)",    
            ylab = "count",
            bins = 100) +
  scale_x_log10(breaks = c(2700, 5400, 12000, 100000, 1000000))
grid.arrange(p1,p2)
```

Zooming in on the higher end, there are around 8,000 donations of \$2,700, and 
still some higher donations esp at \$5,400, but very few that are higher than 
\$12,000. Next I'll look at the individual contributions above $20K.

```{r echo=FALSE, Univariate_Plots_above_20K}
subset(campaign2016, Contribution > 20000)[c(1,2,5,6,7)]

```
There are 10 contributions above $20,000, all occurred in the primaries campaign. 
7 of these are from the "Hillary Victory Fund" and amount to several million 
dollars each - these are clearly outliers. The remaining contributions above 
$20,000 are in the 5 figure region, one to Trump and two to Kasich, both 
occurring long before the primaries started.

Let's look at the impact of the outliers on the mean: 

```{r echo=FALSE, warning=FALSE, Univariate_Plots_summary_with_outliers}
cat("Summary statistics of (all positive) donations :\n")
summary(campaign2016$Contribution)
```



```{r echo=FALSE, warning=FALSE, Univariate_Plots_summary_without_outliers}
cat("Summary statistics after removing Contributions > $1 million  :\n")
summary(subset(campaign2016,Contribution<1000000)$Contribution)
```

Removing outliers above \$1 million reduces the mean from \$124 to \$121 - this 
is still far higher than the median of $28.


```{r echo=FALSE, warning=FALSE, Univariate_Plots_summary_without_0.01percentile}
cat("Summary statistics after removing the top and bottom 0.01% of contributions:\n")
summary(subset(campaign2016, Contribution >= 1 & 
                 Contribution <= 2700)$Contribution)
```

Removing the top and bottom 0.01% quantile reduces the mean further to \$117.2 - 
still far higher than the median of $28. 
These observations imply that, although the outliers of contributions above \$1 
million do significantly raise the mean contribution, the distribution of the 
rest of the contributions is still strongly right-skewed. 
In fact, the strength of the right-skewness is consistent with the notion that 
the contributions are log-normally distributed. To look into the issue of 
whether the bulk of contributions (those between \$1 and \$2,700) is 
log-normally distributed, I will look at their distribution further using 
histograms with a log axis for the amount contributed:


```{r echo=FALSE, warning=FALSE, Univariate_Plots_histogram_upto2700}
qplot(data = subset(campaign2016, Contribution <= 2700 & Contribution >= 1),
      x = Contribution,
      xlab = "Contribution amount ($) (log scale)",
      ylab = "count",bins=100) +
  scale_x_log10()
```

Looking at the bulk of contributions between \$1 and \$2,700 shows the 
discreteness of the donations variable - this is because contributors don't 
arbitrarily choose real numbers but rounded amounts such as \$10, \$20 or \$100 
are far more common.
However, the peaks of the common contributions appears to follow a approximately 
normal distribution in the plot above. The same can be said for bars 
representing less common amounts.
This discreteness can distract from the overall shape of the distribution. For 
that reason I'll study the shape of the distribution further by reducing the 
number of bars in the histogram from 100 to 8 :

```{r echo=FALSE, warning=FALSE, Univariate_Plots_histogram_8bins}
qplot(data = subset(campaign2016, Contribution <= 2700 & Contribution >= 1),
      x = Contribution, 
      xlab = "Contribution (log scale)",
      ylab = "count",
      bins = 8) +
  scale_x_log10(breaks = c(1, 3.16, 10, 31.6, 100, 316, 1000, 2700))
```

When donations between \$1 and $2,700 are represented by a histogram with 8 bins 
and a logarithmic axis for the amount contributed, their distribution comes 
close to a normal distribution - although they are still slightly  right-skewed. 

In what follows, plots involving the variable of the amount contributed will 
often use a logairthmic axis when the approximately log-normal distribution 
matters.

This distribution relates to the *number* of contributions - but how about the 
total sums raised? Small contributions are important but the total sums raised 
are more important from the point of view of election candidates that need to 
run an expensive campaign.
To answer this question, I want to compare the 8 bins in the histogram above to 
the total raised in each of them.

```{r echo=FALSE, warning=FALSE, Univariate_Plots_counts_vs_sums}
#divide [1,2700] into 8 segments (that would be equal-sized on a log axis)
intervals <- c(0,1,2,3,4,5,6,7,8) * log10(2700)/8 
#divide into 8 bins (log) according to the contribution interval
campaign2016$cuts <- cut(campaign2016$Contribution, 
                         breaks = 10^intervals,
                         include.lowest = TRUE)
Contribution_8bins <- subset(campaign2016,Contribution>=1) %>% #ignore <1
  group_by(cuts) %>%  #create df with 8 bins (log-scale)
  summarise(count = n(), #the count of Contributions in bin
             upper_bound = max(Contribution), #the upper bound of the bin
             sum = sum(Contribution),  #total contributions in bin
             median = median(Contribution)) %>%
    arrange(upper_bound)

p1 <- ggplot(data = Contribution_8bins, 
             aes(x = upper_bound, y = count),
             ylab = "count") +
  geom_col() +
  scale_x_log10(breaks = c(1, 10, 100, 1000, 2700))+
  ylab("count")+
  xlab("Contribution (amount up to, $)")+
  #label outliers/above$2700, adapted from https://stackoverflow.com/a/53448372
  geom_label(aes(label = ifelse(upper_bound > 2700, "above $2700", NA)),
             hjust = 1, vjust = 0)+ 
  ggtitle("Number of contributions per segment (histogram)")

p2 <- ggplot(data = Contribution_8bins, aes(x = upper_bound, y = sum/1000000))+
  geom_col()+
  scale_x_log10(breaks = c(1, 10, 100, 1000, 2700))+
  ylab("total (million $)")+
  xlab("Contribution (amount up to, $)")+
  geom_label(aes(label = ifelse(upper_bound > 2700, "above $2700", NA)),
             hjust = 1, vjust = 0)+
  ggtitle("Income from contributions per segment (million $)")
grid.arrange(p1, p2, ncol = 1)

```

In the two plots above, contributions are grouped into 8 bins by amount 
contributed between \$1 and \$2,700, plus a 9th bar for all contributions above 
\$2700 (including outliers). The bin boundaries defining the 8 segments are set 
to fit on a logarithmic axis. The top plot displays the number of contributions 
in each segment - effectively a histogram for the first 8 segments - while the 
bottom plot displays the total income campaigns receive from each of the 
segments (the sum of the contributed amounts in each segment). (The actual 
figures represented in the two plots are shown in the table below.)

Thus, while most common contributions are in the fourth segment (over 3 million 
contributions from \$19.4 to \$52, with a median of \$27), raising only just 
above \$100 million, a much larger source of campaign income -over \$300million- 
is represented by around 125,000 contributions in the 8th segment of \$1005 to 
\$2,700 (median of \$,2700). Noteworthy also that the group of donations above 
\$2700, with a median of \$5400, is the source of $51 million, (i.e. around half 
of the 3 million contributions in the 4 segment) even though it comes from only 
5390 contributions.

The two plots highlight the importance of distinguishing between the 
distribution of the number of contributions and the amounts raised - which are a
reflection of the aproximately log-normal distrubtion of contributions. 


```{r  echo=FALSE, warning=FALSE, Univariate_Plots_counts_vs_sums_table}
Contribution_8bins
```



Let's come back to the negative contribution values - who donates negative money
to a campaign??

```{r echo=FALSE, warning=FALSE, Univariate_Plots_negatives}

dim(returned2016)

```


```{r echo=FALSE, warning=FALSE, Univariate_Plots_3}
summary(returned2016$Contribution)
```

There are around 95 thousand 'negative contributions'. The median and the mean 
are (in absolutely value) larger than the median an the mean of positive 
donations studied above.

For comparison with the previous insights on distribution, I'll look at the 
distribution of contributions in the \$1 - \$2,700 range (using log axis, and showing 
absolute values by multiplying by -1).



```{r echo=FALSE, warning=FALSE, Univariate_Plots4}
p1 <- qplot(x = (-1) * Contribution, data = 
              subset(returned2016, Contribution>=(-2700) & Contribution<=(-1)),
      ylab = "counts", bins = 100) + 
      scale_x_log10() +
      xlab("Contribution ($, absolute value)")
p2 <- qplot(x = (-1) * Contribution, data =
              subset(returned2016, Contribution>=(-2700) & Contribution<=(-1)),
      ylab = "counts", bins=8) + 
      scale_x_log10() +
      xlab("Contribution ($, absolute value)")
grid.arrange(p1, p2, ncol = 1)
```

Negative contributions are skewed towards higher amounts, especially those 
close to $2,700.
 
The original data contain a column for receipt descriptions. What are the most 
common values of this variable?


```{r echo=FALSE, warning=FALSE, Univariate_Plots_receipt_descriptions}
#show the 6 most common receipt description values
sort(table(returned2016$receipt_desc), decreasing = TRUE)[1:6]
#adapted from https://stackoverflow.com/a/17374663
```

"Negative" campaign contributions are typically contributions refunded to donors,
contributions redesignated to the another election (e.g. general), or 
reattributed to the donor's spouse.

Let's look at the 5 number summary for those marked "Reattribution to spouse", 
"Refund", and "redesignation to general" :

```{r echo=FALSE, warning=FALSE, Univariate_Plots_receipt_desc_RTS}
cat("REATTRIBUTION TO SPOUSE  :")
summary(subset(returned2016,receipt_desc=="REATTRIBUTION TO SPOUSE")$Contribution)
```
```{r echo=FALSE, warning=FALSE, Univariate_Plots_receipt_desc_REFUND}
cat("Refund  :")
summary(subset(returned2016,receipt_desc=="Refund")$Contribution)
```
```{r echo=FALSE, warning=FALSE, Univariate_Plots_receipt_desc_RTG}
cat("REDESIGNATION TO GENERAL  :")
summary(subset(returned2016,receipt_desc=="REDESIGNATION TO GENERAL")$Contribution)
```

The median and interquartile range are really rather different from the 
corresponding values for positive donations!

Are these returned donations used equally by all candidates?

```{r echo=FALSE, warning=FALSE, Univariate_Plots_receipt_desc_barplots}
p1 <- ggplot(data=subset(returned2016, receipt_desc=="REATTRIBUTION TO SPOUSE"), 
             aes(CandidateSurname)) +
  geom_bar(stat = "count")+
  xlab("")+
  ggtitle("Marked REATTRIBUTION TO SPOUSE, by candidate")


p2 <- ggplot(data=subset(returned2016, receipt_desc=="REDESIGNATION TO GENERAL"), 
             aes(CandidateSurname)) +
  geom_bar(stat = "count")+
  ggtitle("Marked REDESIGNATION TO GENERAL, by candidate")
grid.arrange(p1, p2, ncol = 1)
```

Some categories of donations are used only by a few top republican candidates,
in particular Cruz.

```{r echo=FALSE, warning=FALSE, Univariate_Plots_receipt_descr_bar_refund}

ggplot(data = subset(returned2016, receipt_desc == "Refund"), 
       aes(CandidateSurname)) +
  geom_bar(stat = "count")+
  xlab("")+
    ggtitle("Marked REFUND, by candidate")+
  coord_flip() 

```

The Refund category is mainly used by the two top Democratic candidates Clinton 
and Sanders.

```{r echo=FALSE, warning=FALSE, Univariate_Plots_top_returned}
subset(returned2016, Contribution < (-15000))[c(1,2,8,9,10)]

```
There are 5 refunded contributions above $15,000.

The shape of negative contributions differs from positive contributions, i.e. 
median and quartiles are much higher.

"Negative" contributions represent money returned to the donor either as a 
refund, or part of being reassigned to a different purpose. While these may not 
seem very interesting, they are commonly the outcome of donors exceeding 
campaign limitations. However, negative values are removed from the main dataset 
and will not studied further.

There are a lot of other variables in the data set which I will be looking at to 
gain a better understanding of the data set. Many are categorical, so often this 
is about looking at counts. First, let's look at election type and party.


```{r echo=FALSE, warning=FALSE, Univariate_Plots_electiontype_party}
p1 <- ggplot(data = campaign2016, aes(Election_Type)) +
  geom_bar(stat = "count") +
  xlab("") +
  ggtitle("Election Type") +
theme(axis.text.x = element_text(angle = 45, hjust = 1)) #from https://stackoverflow.com/a/51147233


p2 <- ggplot(data=campaign2016, aes(Party)) +
  geom_bar(stat = "count") +
    xlab("") +
  ggtitle("Contributions to parties") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(p1, p2, ncol = 2)
```

The majority of contributions were made in the primaries, and only about a third 
of were made in the general election.

The Democratic party received by far the largest number of contributions - over 
twice as many as the Republican party. Contributions to third party candidates 
are barely noticeable.

While most of the remaining variables are categorical, the date of contributions 
can be usefully studied using histograms.


```{r echo=FALSE, warning=FALSE, Univariate_Plots_date}
campaign2016_sample <-sample_n(campaign2016, 73000)

qplot(Date, data = campaign2016_sample, 
      bins = 100, ylab = "Contributions (count)") +
    scale_x_date(date_breaks = "4 month", 
                 limits = as.Date(c('15-01-01','16-31-12')))

```

The histogram of when contributions were made shows a slow rise til early 2016 - 
around February - when contributions reach a much higher level. A peak is 
reached in the week of the election (8. November 2016.). Does this pattern 
correspond to primaries/general election?


```{r echo=FALSE, warning=FALSE, Univariate_Plots_date_primaries_vs_general}
p1 <- qplot(Date, data = subset(campaign2016_sample, Election_Type == "Primary"),
            bins = 100, 
            ylab = "Contributions (count)") +
  scale_x_date(date_breaks = "4 month", 
                 limits = as.Date(c('15-04-01','16-31-12')))+
  #primaries start, adapted from https://stackoverflow.com/a/42237896
    geom_vline(aes(xintercept = as.numeric(as.Date('16-02-01'))), linetype=4) + 
   #final primary
    geom_vline(aes(xintercept = as.numeric(as.Date('16-06-14'))), linetype=4) +
    ggtitle("Primaries")


p2 <- qplot(Date, data = subset(campaign2016_sample, Election_Type == "General"),
            bins = 100, ylab = "Contributions (count)") +
      scale_x_date(date_breaks = "4 month", 
                 limits = as.Date(c('15-04-01','16-31-12'))) + 
   #Republican convention
    geom_vline(aes(xintercept = as.numeric(as.Date('16-07-21'))), linetype=4) +
   #Election day
    geom_vline(aes(xintercept = as.numeric(as.Date('16-11-8'))), linetype=4) +
  ggtitle("General election")

grid.arrange(p1, p2)
```

Vertical lines indicate the start/end of primaries (top) as well as the 
Republican convention (start of general election) and election day (bottom).
Contributions for the general election only really get going once the primaries 
are over.


```{r echo=FALSE, message=FALSE, warning=FALSE, US_states}
#Swing states  Colorado, Florida, Iowa, Michigan, Minnesota, Ohio, Nevada, New 
#Hampshire, North Carolina, Pennsylvania, Virginia, and Wisconsin 
#according to https://en.wikipedia.org/wiki/Swing_state

#SwingStates <- c("CO", "FL", "IA", "MI", "MN", "OH", "NV", "NH", "NC", 
#                 "PA", "VA", "WI")

ggplot(data=campaign2016, aes(State,fill = State)) + #highlight swing states
  geom_bar(stat = "count")+
  scale_fill_manual(values = ifelse(levels(campaign2016$State) %in% SwingStates, 
                                    "red","black"), guide = FALSE )+
         #adapted from https://stackoverflow.com/a/45820586
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

The histogram shows the number of donations per state, swing states are in red. 
The number of contributions is elevated in California, New York and Texas. This 
probably reflects the relative population size.


How about the different candidates? Bar charts of donations grouped by party
affiliation. First Republican candidates.

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate_candidates_byparty_R}
ggplot(data = subset(campaign2016, Party == "Republicans"), 
       aes(CandidateSurname)) +
  geom_bar(stat = "count")+  
  xlab("")+
  ggtitle("Republican candidates") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

There are a lot of Republican candidates, and the number of donations to them 
varies enormously. Some get a lot fewer donations to the extent that they could 
be ignored for analysis, but there is no obvious cut-off that could be used to 
distinguish between the 'main' vs 'marginal' candidates.

Next Democratic Party candidates, and, in one chart, third part candidates, i.e. 
Greens, Libertarians, Independents.

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate_candidates_byparty_D}
p1 <- ggplot(data = subset(campaign2016, Party == "Democrats"), 
             aes(Candidate)) +
  geom_bar(stat = "count")+
  xlab("")+
  ggtitle("Democratic Party candidates")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(data = subset(campaign2016,
                           Party %in% c("Green", "Libertarian", "Independent")), 
             aes(Candidate)) +
  geom_bar(stat = "count")+
  ggtitle("Third party candidates")+
  xlab("")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(p1, p2, ncol = 2)
```

There are much fewer Democratic party candidates, and they clearly separate into 
2 main candidates - Clinton and Sanders - who attract almost all the 
contributions, vs two candidates who get a neglegible number of contributions. 
The number of contributions is far higher than for Republicans, with Sanders 
receiving three times more than the top Republican candidate, Trump.

As for third part candidates, the number of contributions they receive  is not 
in the same league as the two main parties. For the remainder, analysis will 
focus on candidates from the two main parties.


Finally, data is available for the contributor occupation. However, plaintext 
entry means that this is very inconsistent and had to be extensively cleaned. 
However, only the 36 most common occupation categories were used, plus "Info 
Witheld" category. Some of the categories, such as "Retired", "Homemaker" or 
"Disabled" are not jobs, but define demographics whose political participation 
is interesting to study as a group. For the remainder, this variable will 
contain NA:


```{r echo=FALSE, Occupation}
summary(campaign2016$Occupation)
```

While data cleaning categorised a large majority of donors, for 2.4 million 
contributions, the descriptions were too low-quality to easily categorise. 
Let's compare the contributions for which there if occupation information:

```{r echo=FALSE, Univariate_Plots_Occupation}
#exclude NAs
ggplot(data=na.omit(campaign2016), aes(Occupation)) +
  geom_bar(stat = "count")+
  xlab("")+
  ggtitle("Contributor occupation")+
  #scale_y_sqrt()+
  scale_x_discrete(limits = rev(levels(campaign2016$Occupation)))+ #order https://gist.github.com/jennybc/6f3fa527b915b920fdd5
  coord_flip()

```

Contribution occupation is dominated by Retired people, as well as the 
unemployed. To some extent, this is because these category are much wider than 
particular job categories. More specific job categories are dominated by white 
collar jobs, such as lawyers, academics, doctors. Contributions with occupation 
"Info withheld" also stand out.




# Univariate Analysis


### What is the structure of your dataset?

The main dataset (campaign2016) contains information on 7.3 contributions to the 
2016 US presidential election campaign. There are 11 variables with info on the 
amount contributed, the candidate it went to, its date and if it was for the 
primaries or general election, the donor's name, location and occupation. This 
dataset was substantially cleaned, a number of variables have been dropped 
because they are not analysed here. 

In order to drop variables relating to campaign receipt information, negative 
contributions have been split off into a small separate dataframe (returned2016) 
which still contains these variables, and which is analysed briefly early on.

### What is/are the main feature(s) of interest in your dataset?

The campaign contribution amount is the key feature, as it expresses a donor's 
political support and an attempt to influence the election outcome.  To 
interpret this, one needs to study this variable's relationship to other key 
variables, such as which candidate/party the contribution went to, who made the 
contributions, when they were made in the campaign (primary/general, early/late).


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The key feature, the individual election contribution, could be thought of as an
individual vector whose size and direction is an expression of political 
support. To properly investigate this key feature, one needs to look at the 
origin (donor) and target of the vector (candidate/party). One can also consider 
these vectors as being additive in terms of demographics - aggregating the 
contributions of donors sharing by occupation or geographical location can be 
interpreted as an expression of collective political will.

An election campaign is a dynamic process that unfolds over time, and which is 
shaped by external events as well as, for example, some candidates exiting the 
race, the time when contributions were made is important to make sense of what 
is happening.

Candidate party affiliation and election type (primaries/general) are also 
crucial, as during the primaires stage, we're effectively looking at two 
separate races occuring in parallel, before the main "two-horse race" of the 
general election campaign.

Another feature supporting the investigation is occupation: this enables us to 
get an understanding of the different candidates' social base.

There are further features that would be interesting to take into consideration-
in particular, the donor's employer, and their geographical location (ZIP). 
These could be used to study the effect of wider economic intrests, as well as 
demographic effects (urban vs rural, areas of deprivation, etc) using additional
socio-economic data. However, the work required is beyond the scope of the 
present study.


### Did you create any new variables from existing variables in the dataset?

Yes: the candidates' party affiliation was not originally available, and needed 
to be created as a variable. 

The very low-quality occupation data was used to create a new variable with 37 
different categories.

Also, Candidate surname was created as the full-name often takes up too much 
space in axis labels.

### Of the features you investigated, were there any unusual distributions? 

Yes - intersting features of the contribution amount are the outliers, the 
log-normal distribution, and the negative contributions.

The extreme outliers of contributions amounting to several million dollars to 
the Clinton campaign in fact come from the Hillary Victory Fund. These outliers 
- well beyond the normal contribution limitations - in fact represent an 
aggregation of much smaller donations. However this joint fund was 
controversial, with Sander's alleging it aimed to "skirt fundraising limits on 
[Clinton's] presidential campaign". 
(source: https://en.wikipedia.org/wiki/Hillary_Victory_Fund)

Looking at quantiles, the 99.9 percentile is \$2700 ; while the 99.99 percentile 
is \$5400. Interestingly these values correspond to legal limits on 
contributions (donations limit for individuals and married couples, 
respectively) and thus reflect how regulatory measures shape how campaign 
finance works. 
(source: https://transition.fec.gov/pages/brochures/citizens.shtml)

After removing outliers, and restricting the analysis to contributions ranging 
from \$1 to \$2700, it became clear that its distribution approximates a 
log-normal distribution (though this is still slightly right-skewed). This 
distribution probably reflects the wealth distribution in US society - which is 
also log-normal.

Finally, there are negative contributions whose distribution varies from the rest of contributions. These contributions are often associated with campaign finance limits. A common approach when a donation exceeds the legal individual limit of $2,700 is to re-attribute the excess to a donor's spouse - which is then registered as a negative contribution.

### Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Yes, the low-quality occupation data was reduced and factorised to 37 different 
categories - though over 2 million contributions resisted this data cleaning 
measure. This data cleaning measure simultaneously made data analysis possible 
as well as reducing the memory footprint of this variable.

Also, a number of records were dropped - those not relating to either the 
primaries or general elections. "Negative donations" were split off into a 
separate dataframe in which variables only relevant to the analysis of negative 
donations were preserved.


# Bivariate Plots Section
The variables investigated in this section :
* date vs election type variables: comparison of the primaries and general 
election over time
* candidate vs contribution amount variables: comparison of candidates of the 
same party according to the contribution amount. Since only a single candidate 
per party remains in the general election, the data are restricted to donations 
in the primaries
* candidate vs date variables : comparison of condidates in terms of when they 
received donations
* party/candidate vs contribution amount: relative aggregated ontributions and 
preference for one candidate over another

To compare primaries and general election over the contribution date, a 
frequency polgon is a good way of comparing these, and using log y-axis given 
the log-normal distribution of contributed amounts.

```{r echo=FALSE, message=FALSE, warning=FALSE, BivPlots_GenvsPrim_freqpoly}
qplot(Date, data = subset(campaign2016), 
      binwidth=7,
      geom='freqpoly',
      color=Election_Type,
      ylab="Contributions per week (count) (log)")+
  scale_x_date(date_breaks = "4 month", 
                 limits = as.Date(c('15-01-01','16-31-12')))+
  scale_y_log10()

```

The (total) number of contributions per week continues to rise over the course 
of the campaign. However there is a clear cross-over point when the primaries 
finish. Interestingly, general election contributions do occur during the 
primaries, and primaries contributions are recorded during the general election 
campaign - the latter may be candidates that are no longer in the race but still 
need to pay off expenses.  During the primaries, primaries contributions are 
around 2 orders of magnitude (100fold) larger than those for the general 
election until the cross-over point in August 2016.

What about the total amounts contributed during the primaries and general 
election campaigns?

```{r echo=FALSE, message=FALSE, warning=FALSE, BivPlots_GenvsPrim_daily_total}
ggplot(data = campaign2016, aes(y = Contribution, 
                                x = Date, colour = Election_Type))+
  geom_line(stat = "summary", fun.y = sum)+
  ylab("Total sum contributed per day")+
  scale_x_date(date_breaks = "4 month", 
                 limits = as.Date(c('15-01-01','16-31-12')))
```

The sum of daily contributions follows a similar pattern as the number of 
contributions in that there is a clear cross-over point in August 2016. However, 
the timeline for the sum of contirbutions is noisier as it is daily rather than 
the weekly number of contributions shown in the previous plot. The total amounts 
are likely also influenced by the time points at which specific candidates leave 
the race. The visible spikes are due to the outliers identified in the earlier 
section.


Now for a closer look at the primaries and what happens between candidates. 
During the primaries, the race is between multiple candidates of the same 
parties, unlike in the general election, when it is between different parties. 
I will starting by looking at the Democrats.

```{r echo=FALSE, message=FALSE, warning=FALSE, BivPlots_DemsPrim_boxplot}
qplot(y = Contribution, x = Candidate, 
      data = subset(campaign2016, 
                    Candidate %in% Democrats & Election_Type == "Primary"),
      geom='boxplot', ylab = "Contribution amount ($) (log scale)")+
  stat_summary(fun.y = mean, geom='point',shape=4) + #add cross for mean
  scale_y_log10()
```

The mean contribution for each candidate is indicated by a cross.
There are quite substantial differences in the average donations for different 
candidates, with Sanders and  Clinton significantly lower median contributions 
than the other democratic candidates. There are quite prominent outliers for 
Clinton of several million dollars - these are the outliers previously 
discussed. For Clinton, the mean contribution (cross) is higher than the median 
- this is due to these outliers. In comparison, for Lessig, the mean is lower 
than the mean, due to a number of outliers at the low end.

To look more closely at the mean and five number summaryfor the two main 
Democratic party candidates, here are the summary stats:

Clinton summary statistics:
```{r echo=FALSE, Bivariate_Plots_DemsPrim_Clinton_summary}
summary(subset(campaign2016,
               CandidateSurname == "Clinton" & 
                 Election_Type  == "Primary")$Contribution)

```
Sanders summary statistics:
```{r echo=FALSE, Bivariate_Plots_DemsPrim_Sanders_summary}
summary(subset(campaign2016, 
               CandidateSurname == "Sanders" & 
                 Election_Type  == "Primary")$Contribution)

```

At \$180, the mean contribution in the primaries for Clinton is much higher than 
for Sanders (\$47.49). Is this entirely due to the outliers? Here are the 
summary statistics for Clinton, with outliers removed by only considering 
contributions up to \$10.000 (the maximum for Sanders):

```{r echo=FALSE, Bivariate_Plots_DemsPrim_Clinton_summary2}
summary(subset(campaign2016,
               CandidateSurname == "Clinton" & 
                 Election_Type=="Primary" & 
                 Contribution < 10000)$Contribution)
```

Removing the outliers lowers the mean contribution for Clinton in the primaries 
to \$162.41 - but this is still much higher than for Sanders, even though the 
median amount contributed is relatively similar (\$25 for Clinton, \$27 for 
Sanders). 
This means that, while the distribution of contributions is right-skewed for 
both candidates, it is far more skewed for Clinton, and on the basis of these 
numbers, the expectation is that a higher proportion of contributions received 
by Clinton fall in the higher categories.

To compare the distribution of contributed amounts between Clinton and Sanders, 
we can plot a histogram of the contribution amounts (to compare the dstributions 
for the two candidates, the amounts are normalised by calculating the density, 
and only contributions between \$1 and \$10.000 are considered):

```{r echo=FALSE, Bivariate_Plots_DemsPrim_ClintonvsSanders_density}
#adapted from https://stackoverflow.com/a/22181949
ggplot(aes( x = Contribution, fill = CandidateSurname), 
       data = subset(campaign2016,
                     Contribution>1 & 
                       Contribution <10000 & 
                       CandidateSurname %in% c("Clinton","Sanders") & 
                       Election_Type=="Primary")) +
  geom_histogram(aes(y = ..density..),    
                 alpha = 0.5,
                 position ='identity',
                 bins = 10) +
  scale_x_log10() +
  xlab("Contribution amount ($)")

```

Amounts are plotted on a log-axis and 10 bins are used. Looking at these 
"log-transformed deciles", a higher proportion of the contributions to the 
Sanders campaign fall in the 4th and 5th "log-decile", while the Clinton primary 
campaign receives a higher proportion of contributions in the 6., 7., 8., 
9. "log-decile".

When did Democratic party candidates receive the most donations? We can look at 
this using a frequency polygon of the number of donations received each week. 
The vertical line marks the first primary contest on February 1, 2016. While 
Lessig and Webb had suspended their campaigns before the start of the first 
primary, O'Malley withdrew following 1. February, after which Clinton and 
Sanders were the only 2 candidates left in the contest.  The second vertical 
line marks the final primary in the District of Columbia on June 14.
(source: 
https://en.wikipedia.org/wiki/2016_Democratic_Party_presidential_primaries)

```{r echo=FALSE, warning=FALSE, Bivariate_Plots_DemsPrim_freqpoly}
qplot(Date, data = subset(campaign2016, Party == 'Democrats' & 
                            Election_Type == 'Primary'),
      binwidth = 7,
      geom = 'freqpoly',
      color = Candidate,
      ylab = "Contributions per week")+
    scale_x_date(date_breaks = "4 month", 
                 limits = as.Date(c('15-04-01', '16-08-01')))+
   #add vertical lines, adapted from https://stackoverflow.com/a/42237896
  geom_vline(aes(xintercept = as.numeric(as.Date('16-02-01'))), linetype=4) +
  geom_vline(aes(xintercept = as.numeric(as.Date('16-06-14'))), linetype=4)

```

Using a linear rather than a log axis, the contributions for any candidates 
other than Clinton and Sanders are unnoticeable.

While Sanders' contributions peaked in the middle of the primaries - perhaps as 
a result of enthusiasm for Sanders ebbing as his chances of winning the 
primaires waned, Clinton campaign contributions continued to rise as the general 
election approached. 

How does this relate to the total amount of contributions received?

```{r echo=FALSE, Bivariate_Plots_DemsPrim_nb_vs_total}
p1 <- ggplot(aes(CandidateSurname),
             data = subset(campaign2016,
                           Party == "Democrats" & Election_Type == 'Primary')) +
  geom_bar() + #count of Contributions
    ggtitle("Nb contributions to candidates")

p2 <- ggplot(aes(CandidateSurname),
             data = subset(campaign2016, 
                           Party == "Democrats" & Election_Type == 'Primary')) +
  geom_bar(aes(weight = Contribution/1000000)) + 
              #weight parameter calculates the sum of Contributions rather than
              #count https://ggplot2.tidyverse.org/reference/geom_bar.html
    ggtitle("Total ($) contributed to candidates") +
  ylab("sum (million $)")
grid.arrange(p1, p2, ncol = 2)
```

Two things stand out here: (1) candidates other than Clinton and Sanders are 
almost invisible ; (2) while Sanders receives far more contributions than 
Clinton, Clinton receives more than twice as much money. In part this is due to 
the fact that the "Hilary Clinton Victory fund" aggregates numerous smaller 
donations. However, as point out above, even if we exclude these, Clinton's 
campaign contributions are much more right-skewed as she receives more large 
donations than Sanders.

Let's look at a new variable: donor occupation. Who are the people donating 
money to the two main Democratic party candidates? Which are the top 5 
categories for each candidate?  (totals in million dollars)

```{r echo=FALSE, message=FALSE, warning=FALSE, BivPlots_Dem_Pri_byOccupation}

#group, calculate n, median,mean, sum, rel stdev
byjob.DemPrim <- subset(campaign2016,
                        Election_Type=="Primary" & 
                          CandidateSurname %in% c("Sanders","Clinton")) %>%
  group_by(Occupation,CandidateSurname) %>%
  summarise(n = n(),
            Total = sum(Contribution)/10^6, #amount in $million
            Mean = mean(Contribution),
            Median = median(Contribution)) %>%
  ungroup()


byjob.T <- dcast(na.omit(byjob.DemPrim), 
                 Occupation ~ CandidateSurname, value.var = "Total")


#sort by total amount going to Sanders, descending
byjob.T <- arrange(byjob.T,-Sanders)

p1 <- ggplot(aes(x = reorder(Occupation,Sanders), y = Sanders),
       data = byjob.T[1:5,])+ #https://stackoverflow.com/a/6901597
       geom_bar(stat="identity") +
       #scale_y_sqrt()+
      coord_flip()

#sort by total amount going to Clinton, descending
byjob.T <- arrange(byjob.T,-Clinton)

p2 <- ggplot(aes(x = reorder(Occupation,Clinton), y = Clinton),
       data = byjob.T[1:5,]) +
       geom_bar(stat="identity") +
       #scale_y_sqrt()+
      coord_flip()

grid.arrange(p1, p2, ncol=1)

```

The difference betwen the two candidates is striking. While Retired people and 
those who withold occupation info are to be expected to be in the top 5 because 
they are the largest catgories overall, Sanders receives far more money from 
unemployed people.

To some extent, this ranking will be determined by how occupation categories 
were defined during the data cleaning step. To make this more meaningful as a 
comparison between the support base of the two candidates, I want to look at the 
number of contributions (to remove the effect of some demographics having much 
higher disposable income) and at the imbalance in which candidate is most 
preferred by donors of the same occupation.


```{r echo=FALSE, Bivariate_Plots_Dem_Pri_byOccupation_preference}

byjob.n <- dcast(byjob.DemPrim , Occupation ~ CandidateSurname, value.var = "n")
byjob.n$ratio <- byjob.n$Clinton/byjob.n$Sanders
byjob.n <- arrange(byjob.n,ratio)

 p1 <- ggplot(aes(x = reorder(Occupation,1/ratio), 
                  y = 1/ratio),
              data = byjob.n[1:6,]) + #https://stackoverflow.com/a/6901597
       geom_bar(stat = "identity") +
       scale_y_sqrt() +
      coord_flip() +
     ggtitle("Preference for Sanders")


 p2 <- ggplot(aes(x = reorder(Occupation,ratio), 
                  y = ratio),
              data = byjob.n[33:38,]) + #https://stackoverflow.com/a/6901597
       geom_bar(stat = "identity") +
       scale_y_sqrt()+
      coord_flip() +
        ggtitle("Preference for Clinton")


grid.arrange(p1, p2, ncol=1)

```

Self-employed donors are 80 times more likely to contribute money to Sanders 
than to Clinton. Unemployed donors are over 20 times more  likely to contribute 
money to Sanders. Sanders is also much preferred by donors in someblue-collor 
occupations - eletcricians, mechanics, truck drivers.

By comparison, those occupations which favour Clinton over Sanders only express 
a milder bias of maximum 4 times.


Next for a similar analysis of Republican primaries. 

```{r echo=FALSE, Bivariate_Plots_Republican_Primaries}
ggplot(aes(Candidate), data = subset(campaign2016, Party == "Republicans" &  
                                       Election_Type == 'Primary'), 
       ylab = "sum($)") +
  geom_bar() +
  ggtitle("Contributions to candidates in Republican primaries") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

There were 17 candidates from the Republican party. As it can be difficult to 
visualise this many candidates in a plot, I will focus more on the final 
5 which remained in the race. What are the total amounts received by each 
candidate's campaign?

```{r echo=FALSE, Bivariate_Plots_Rep5Prim_Totals_barplot}
Rep_final5 <- c("Trump", "Kasich", "Cruz", "Rubio", "Carson")
ggplot(aes(CandidateSurname), data = subset(campaign2016,
                                            Election_Type == 'Primary' & 
                                            CandidateSurname %in% Rep_final5)) +
  geom_bar(aes(weight = Contribution/1000000)) + 
  ggtitle("Total ($) donated to Republican candidates in primaries") +
  ylab("Amount contributed (million $)")
  #theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Contributions to the primary campaign of Cruz was significantly larger than 
Trump, Rubio or Carson. However, it is important to note that we're only looking 
at registered contributions data, and billionaire candidate such as Trump 
engaged in significant amount of self-financing (\$66 million).

Also, with conribution totals ranging from around \$10 to over \$60 million for 
the top 5 Republican candidates, the primary campaigns  of the final 5 
candidates received much less than the two final Democrat candidates.
(source: https://www.opensecrets.org/pres16/candidate?id=N00023864)

How much did the remaining Republican candidates receice?

```{r  echo=FALSE, Bivariate_Plots_RepOtherPrim_Totals_barplot}

Rep_others=c("Christie","Bush","Fiorina","Gilmore","Graham","Huckabee",
             "Jindal","Johnson","Pataki","Paul","Perry","Santorum","Walker")

ggplot(aes(CandidateSurname), 
       data = subset(campaign2016,
                     Election_Type == 'Primary' & 
                       CandidateSurname %in% Rep_others)) +
  geom_bar(aes(weight = Contribution/1000000)) + 
  ylab("Amount contributed (million $)") +
  ggtitle("Total ($) donated to Republican candidates in primaries") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

Looking at the other Republican candidates, it is noticeable how much less money 
they attracted in comparison to the two main Democratic party candidates. Could 
it be that this is the result of the much larger number of candidates? I.e. is 
the same amount of funding shared out between a larger number of candidates?


```{r  echo=FALSE, Bivariate_Plots_Rep_vs_Dem_primaries}
p1 <- ggplot(aes(Party),
             data = subset(campaign2016,
                           Election_Type == 'Primary' & 
                             Party %in% c("Democrats","Republicans"))) +
  geom_bar(aes(weight = Contribution/1000000)) + 
  ggtitle("Contributions ($) in primaries")+
  ylab("Amount contributed (million $)")

p2 <- ggplot(aes(Party),
             data = subset(campaign2016,
                           Election_Type == 'Primary' & 
                             Party %in% c("Democrats","Republicans"))) +
  geom_bar() + 
  ggtitle("Contributions (#) in primaries") +
  ylab("Contributions (count)")

grid.arrange(p1, p2, ncol = 2)

```
The total amount of contributions received by Republican party candidates was 
somewhat smaller - perhaps 1/5 - and shared between 17 different candidates. 
There are also less than half the number of contribution to the Republican 
candidates in the primaries.

Now to continue with the analysis of the final 5 Repulican candidates with a box plot.
```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots_Rep5Prim_boxplot}
qplot(y=Contribution, x = CandidateSurname,
      data = subset(campaign2016,
                    CandidateSurname %in% Rep_final5 & 
                      Election_Type=="Primary"),
      geom = 'boxplot', ylab = "Contribution ($) (log scale)") +
  stat_summary(fun.y = mean, 
               geom='point',
               shape=4) + #add cross for mean
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

The median contributed amount was highest for the Kasich and Rubio campaigns. 
Trump received the lowest average (median) contributions of the final 5 
candidates - but also the highest outliers.

```{r echo=FALSE, warning=FALSE, Bivariate_Plots_RepsPrim_freqpoly}
qplot(Date, 
      data = subset(campaign2016, 
                    CandidateSurname %in% Rep_final5 & 
                            Election_Type == 'Primary'),
      binwidth = 7,
      geom = 'freqpoly',
      color = Candidate,
      ylab = "Contributions per week")+
    scale_x_date(date_breaks = "4 month", 
                 limits = as.Date(c('15-04-01', '16-08-01'))) +
  #primaries start
  geom_vline(aes(xintercept = as.numeric(as.Date('16-02-01'))), linetype=4)+ 
  #Cruz, Kasich withdraw
  geom_vline(aes(xintercept = as.numeric(as.Date('16-05-03'))), linetype=3)+
  #primaries end
  geom_vline(aes(xintercept = as.numeric(as.Date('16-06-07'))), linetype=4) 

```


The start and the end of the primaries contests are shown by vertical lines.
Cruz and Kasich suspended their campaigns after the Indiana primaries on 3.May 
(middle line), after which Trump remained as the sole candidate. Carson withdrew 
on 4.March, Rubio on 15. March.

(source: 
https://en.wikipedia.org/wiki/2016_Republican_Party_presidential_primaries)

Surprisingly perhaps, Trump received the vast majority of primary contributions 
after the end of the primaries. This is quite different from the other 
candidates who received the bulk of contributions during their primary campaign. 


```{r echo=FALSE, message=FALSE, warning=FALSE, BivPlots_Rep_Pri_byOccupation}

#group, calculate n, median,mean, sum, 
byjob.RepPrim <- subset(campaign2016,
                        Election_Type=="Primary" & 
                          CandidateSurname %in% c("Trump","Cruz")) %>%
  group_by(Occupation,CandidateSurname) %>%
  summarise(n = n(),
            Total = sum(Contribution)/10^6, #amount in $million
            Mean = mean(Contribution),
            Median = median(Contribution)) %>%
  ungroup()


byjob.T <- dcast(na.omit(byjob.RepPrim), 
                 Occupation ~ CandidateSurname, value.var = "Total")


#sort by total amount going to Trump, descending
byjob.T <- arrange(byjob.T,-Trump)

p1 <- ggplot(aes(x = reorder(Occupation,Trump), y = Trump),
       data = byjob.T[1:5,])+ #https://stackoverflow.com/a/6901597
       geom_bar(stat="identity") +
       #scale_y_sqrt()+
      coord_flip()

#sort by total amount going to Cruz, descending
byjob.T <- arrange(byjob.T,-Cruz)

p2 <- ggplot(aes(x = reorder(Occupation,Cruz), y = Cruz),
       data = byjob.T[1:5,]) +
       geom_bar(stat="identity") +
       #scale_y_sqrt()+
      coord_flip()

grid.arrange(p1, p2, ncol=1)

```

Compared to the Democrats, the two main Republican candidates have a more 
similar donor base.

```{r echo=FALSE, Bivariate_Plots_Rep_Pri_byOccupation_preference}

byjob.n <- dcast(byjob.RepPrim , Occupation ~ CandidateSurname, value.var = "n")
byjob.n$ratio <- byjob.n$Cruz/byjob.n$Trump
byjob.n <- arrange(byjob.n,ratio)

 p1 <- ggplot(aes(x = reorder(Occupation,1/ratio), 
                  y = 1/ratio),
              data = byjob.n[1:6,]) + #https://stackoverflow.com/a/6901597
       geom_bar(stat = "identity") +
       scale_y_sqrt() +
      coord_flip()+
        ggtitle("Preference for Trump")


 p2 <- ggplot(aes(x = reorder(Occupation,ratio), 
                  y = ratio),
              data = byjob.n[33:38,]) + #https://stackoverflow.com/a/6901597
       geom_bar(stat = "identity") +
       scale_y_sqrt()+
      coord_flip() +
        ggtitle("Preference for Cruz")


grid.arrange(p1, p2, ncol=1)

```

Again, compared to the Democrats, the preferences for either of the two top 
Republican candidates are less pronounced.


Now I want to analyse the general election campaign using similar plots. First 
let's look at the general election campaign donations of the different 
candidates.

```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots_General_boxplot}
qplot(y = Contribution, x = Candidate,
      data = subset(campaign2016,
                    CandidateSurname %in% c("Clinton","Trump","Stein",
                                            "McMullin","Johnson") & 
                      Election_Type=="General"),
      geom = 'boxplot', 
      ylab = "Contribution ($) (log scale)") +
  stat_summary(fun.y = mean, 
               geom = 'point',
               shape = 4) + #add cross for mean
  scale_y_log10()

```

Perhaps surprisingly given observations for the primaries, Clinton received the 
lowest mean and median contributions of all the candidates in the general 
election campaign. Third party candidates and Trump were similar.


```{r echo=FALSE, Bivariate_Plots_General_Clinton_summary}
summary(subset(campaign2016, 
               CandidateSurname == "Clinton" & 
                 Election_Type == "General")$Contribution)
```


```{r echo=FALSE, Bivariate_Plots_General_Trump_summary}
summary(subset(campaign2016, 
               CandidateSurname =="Trump" & 
                 Election_Type=="General")$Contribution)
```

Across all quartiles, general election contributions received by Trump are much 
higher than those received by Clinton - with the mean being over twice as high 
for Trump.

Does this disparity in the contribution amount extend to the number of 
contributions?

```{r echo=FALSE, Bivariate_Plots_General_Clinton_vs_Trump_freqpoly}
qplot(Date, data = subset(campaign2016,
                          CandidateSurname %in% c("Clinton","Trump") &
                            Election_Type=="General"),
      binwidth = 7,
      geom = 'freqpoly',
      color = Candidate,
      ylab = "Contributions per week")+
    scale_x_date(date_breaks = "4 month")+
  geom_vline(aes(xintercept = as.numeric(as.Date('16-07-20'))), linetype=4) +
  geom_vline(aes(xintercept = as.numeric(as.Date('16-11-08'))), linetype=4) 

```

Vertical lines show the start and end of the general election campaigns.

Looking at the number of weekly contributions, the Clinton campaign is 
consistently and overwhelmingly more popular during the general election 
campaign.


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

It's important to distinguish between the number of contributions and the summed 
amount as these are often at odds. For example, while Sanders received far more 
contributions in the primaries, the the toal sum raised by Clinton is far higher.  
Similarly, while Democratic Party candidates received over wice as many donations 
than Republicna party candidates in the primaries, the sums raised are much 
closer together. What this observation means is that some candidates and parties 
are preferred by donors making large contributions.

Another interesting observation is that the total sums raised in the primaries 
by the main Republican candidates is significantly smaller those of the main 
Democratic candidates, probably because the number of Republican candidates was 
much larger. Plausibly, the total amount of finance that can potentially be 
raised in each party's primaries is limited, so a larger number of candidates 
has this effect.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Analaysis of the contribution date reveals some interesting relationships: while 
most candidates in the primaries received the bulk of donations during the 
primaries contest as well as in the runup to the primaries, Trump received the 
majority of primaries donations after the primaries had finished. 

As for the date of donations to the general election campaign, while the stream 
of donations to Clinton increased throughout the period of the general election, 
donations to the Trump general election peaked early on in the general election 
campaign, and then were at a relatively low level.


### What was the strongest relationship you found?

Comparing the support base of Clinton and Sanders using the occupation variable 
was instructive: some demographics had a very strong preference in who they 
donated to. In particular self-employed and unemployed donors were 80 times and
over 20 times more likely to donate to Sanders. This may be indicative of 
Sanders mobilising different demographics.


# Multivariate Plots Section

The first relationship I want to get a better overview of is a comparison of 
candidates during the primaries race in terms of the relative density 
distribution of contributions over time - i.e. the variables of candidate, date, 
number of contributions restricted by election type.

In order to do this and capture how the primary race within a party unfolds, I 
will use violin plots to look at when candidates received most of the 
contributions. In volin plots, the relative density of contributions is 
visualised, i.e. contributions are normalised for each candidate. These will use
Date as the x-variable, with Candidate as the third ctegorical variable, 
restricted to the primary election campaigns.

```{r echo=FALSE, Multivariate_Plots_DemsPrim_violin}
ggplot(aes(x = Candidate, y = Date),
       data = subset(campaign2016,
                     Party == "Democrats" & Election_Type == 'Primary')) +
         geom_violin() +
      scale_y_date(date_breaks = "4 month") +
  geom_hline(aes(yintercept = as.numeric(as.Date('16-02-01'))), linetype=4)+
  geom_hline(aes(yintercept = as.numeric(as.Date('16-06-14'))), linetype=4)+
  coord_flip() # time on horizontal axis

```

The violin plot shows the distribution in time of all of a campaign's 
contributions, with the date on the horizontal axis.

Webb and Lessig received most contributions long before the primaries started. 
O'Malley received the bulk of contributions before 1. February, and while they 
markedly declined after he withdrew from the race, contributions continued for 
several months. It is not unusual to still receive donations after a candidate 
withdraws from the race:
"In US elections, suspending a campaign allows candidates to cease active 
campaigning while still legally raising funds to pay off their debts" - source: https://en.wikipedia.org/wiki/2016_Democratic_Party_presidential_primaries

The evolution of contributions is relatively different for the other two 
candidates. Sanders campaign contributions peaked in May, while for the 
Clinton primary campaign, the number of contributions kept increasing until the 
end of the primaries.

Here is the equivalent violin plot for the top 5 Republican candidates:

```{r echo=FALSE, Multivariate_Plots_Rep5Prim_violin}
ggplot(aes(x = CandidateSurname, y = Date),
       data = subset(campaign2016, 
                     CandidateSurname %in% Rep_final5 &
                       Election_Type == 'Primary')) +
         geom_violin() +
      scale_y_date(date_breaks = "4 month")+
   #primaries start
  geom_hline(aes(yintercept = as.numeric(as.Date('16-02-01'))), linetype=4) +
  #Cruz, Kasich withdraw
  geom_hline(aes(yintercept = as.numeric(as.Date('16-05-03'))), linetype=3) +
   #primaries end
  geom_hline(aes(yintercept = as.numeric(as.Date('16-06-07'))), linetype=4) +
  coord_flip() # time on horizontal axis

```

The start and the end of the primaries contests are shown by vertical lines. 
Cruz and Kasich suspended their campaigns after the Indiana primaries on 3.May 
(middle line), after which Trump remained as the sole candidate.

Surprisingly perhaps, Trump received the vast majority of primary contributions 
after the end of the primaries. This is quite different from the other 
candidates who received the bulk of contributions during their primary campaign. 
Also noticeable how much earlier Rubio started to receive campaign funding.
Carson withdrew on 4.March, Rubio on 15. March.

Below is the correpsonding violon plot for the two main candidates in the 
general election:

```{r echo=FALSE, Multivariate_Plots_General_Clinton_vs_Trump_violin}
ggplot(aes(x = Candidate, y = Date),
       data = subset(campaign2016, 
                     CandidateSurname %in% c("Clinton","Trump") & 
                       Election_Type=="General")) +
         geom_violin() +
      scale_y_date(date_breaks = "4 month") +
  #add lines for start/end of primaries/party conventions
  #geom_hline(aes(yintercept = as.numeric(as.Date('16-02-01'))), linetype=4)+
  #geom_hline(aes(yintercept = as.numeric(as.Date('16-05-03'))), linetype=3)+
  coord_flip() # time on horizontal axis

```

The previous analysis looked at the occupations of supporters of the different 
candidates, however this was limited by the number of occupation categories that 
could be investigated, and the metric (number of donations vs sums raised). 
Treemaps are an excellent multivariate type of plot to get a more complete 
understanding of different candidates' support base. In the plots below, the 
size of rectangles is an expression of the number of contributions, with the 
colour hue encoding the median contribution donated by people with the same 
occupation.

First a treemap of the relative support bases of Clinton and Sanders during 
the primaries.

```{r echo=FALSE, warning=FALSE, Multivariate_Plots_ClintonSander_tm}
#use previously calculated dataframe, omitting NA
byjob.DemPrim <- na.omit(byjob.DemPrim)
ggplot(subset(byjob.DemPrim), 
              aes(area = Total,  
           fill = Median, 
           subgroup = CandidateSurname, 
           label = Occupation)) +
  geom_treemap() +
    geom_treemap_subgroup_border() +
  geom_treemap_subgroup_text(place = "centre", 
                             grow = T, 
                             alpha = 0.5, 
                             colour = "yellow", 
                             fontface = "italic", 
                             min.size = 0) +
  geom_treemap_text(
    #fontface = "italic",
    colour = "white",
    place = "left",
    grow = TRUE
  )

```

While the occupations that most strongly support Sanders - the unemployed, 
programmers- differ from Clinton, many of the remaining groups are quite 
similar. However there are some drastic differences in the total of 
contributions - for example lawyers and senior managers donate far more money 
to Clinton. It is also intersting that the group with the highest median 
donations is people witholding information on their occupation.

How about support for the top Republican candidates in the primaries?

```{r echo=FALSE, warning=FALSE, Multivariate_Plots_Reps_tm}
#group, calculate n, median,mean, sum, 
byjob.RepPrim <- subset(campaign2016, 
                        Election_Type=="Primary" & 
                          CandidateSurname %in% c("Cruz","Trump",
                                                  "Carson","Rubio")) %>% 
  group_by(Occupation,CandidateSurname) %>%
  summarise(n = n(),
            Total = sum(Contribution),
            Mean = mean(Contribution),
            Median = median(Contribution)) %>%
  ungroup()

ggplot(subset(na.omit(byjob.RepPrim)), 
       aes(area = Total,  
           fill = Median, 
           subgroup = CandidateSurname, 
           label = Occupation)) +
  geom_treemap() +
    geom_treemap_subgroup_border() +
  geom_treemap_subgroup_text(place = "centre", 
                             grow = T, 
                             alpha = 0.5, 
                             colour = "yellow", 
                             fontface = "italic", 
                             min.size = 0) +
  geom_treemap_text(
    #fontface = "italic",
    colour = "white",
    place = "left",
    grow = TRUE
  )

```

The support base for the four top Republican canidates is relatively similar, 
with Retired people playing a large role. Some patterns are discernible - 
business owners prefer Trump, homemakers prefer Cruz- but there doesn't seem to 
be a particular demographic being mobilised unexpectedly (as can be seen with 
Sanders and unemployed donors)

Let's use the same analysis to compare Trump and clinton in the general election:

```{r echo=FALSE, warning=FALSE, Multivariate_Plots_ClintonTrump_tm}
#group, calculate n, median,mean, sum
byjob <- subset(campaign2016, Election_Type == "General" & 
                  CandidateSurname %in% c("Trump","Clinton")) %>% 
  group_by(Occupation,CandidateSurname) %>%
  summarise(n = n(),
            Total = sum(Contribution),
            Mean = mean(Contribution),
            Median = median(Contribution)) %>%

  ungroup()
byjob <- na.omit(byjob)
ggplot(subset(byjob), aes(area = Total, 
                          fill = Median, 
                          subgroup = CandidateSurname, 
                          label = Occupation)) +
  geom_treemap() +
    geom_treemap_subgroup_border() +
  geom_treemap_subgroup_text(place = "centre", grow = T, alpha = 0.5, colour =
                             "yellow", fontface = "italic", min.size = 0) +
  geom_treemap_text(
    #fontface = "italic",
    colour = "white",
    place = "left",
    grow = TRUE
  )

```

3 things stand out: (1) Clinton receives significantly more money during the 
general election (2) while some groups are shared, others have an aversion to 
Trump - e.g. teachers, academics, the unemployed, those working in the arts ; 
(3) groups that donate to both candidates donate much higher amounts to Trump - 
senior managers, doctors, lawyers. 

Alluvial diagrams are another multivariate plot suitable to gaining insight into 
the multifaceted flows of campaign finance. Below is an alluvial diagram of 
contributions coming from retired donors (in million $).

```{r echo=FALSE, Multivariate_Plots_Retired_alluvial}

retired <- subset(campaign2016,Occupation == "Retired") %>%
    group_by(Election_Type,Party,CandidateSurname) %>%
    summarise(n = n(),
            Total = sum(Contribution)/10^6,
            Mean = mean(Contribution),
            Median = median(Contribution)) %>%
  arrange(Party) %>%
  ungroup()

#only consider if above 6000 contributions
alluvial_plot <- filter(retired, n>6000 &
                          ((Election_Type == "General" & 
                              CandidateSurname %in% c("Trump","Clinton")) | 
                             (Election_Type == "Primary")))


ggplot(alluvial_plot,
       aes(y = Total, axis3 = Election_Type, axis1 = Party, 
           axis2 = CandidateSurname)) +
  geom_alluvium(aes(fill=Party), width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label_repel(stat = "stratum", label.strata = TRUE,size=3)+
  scale_x_discrete(limits = c("Party", "Candidate", "Type"), 
                   expand = c(.05, .05)) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  ggtitle("Campaign donations from Retired Contributors")


```

Retired donors contribute a large amount of campaign finance to both of the main 
parties. The majority of it is donated in the primaries. Sanders received far 
less money from retired contributors than Clinton. Trump was the main recipient 
for Republican candidates, but Clinton received significantly more than Trump, 
in both the primaries and the general election.

Another demographic mentioned before are the unemployed. Which candidates did 
they donate their money to?

```{r echo=FALSE, Multivariate_Plots_Unemployed_alluvial}
unemployed <- subset(campaign2016,Occupation == "Unemployed") %>%
    group_by(Election_Type,Party,CandidateSurname) %>%
    summarise(n = n(),
            Total = sum(Contribution)/10^6,
            Mean = mean(Contribution),
            Median = median(Contribution)) %>%
  arrange(Party) %>%
  ungroup()

alluvial_plot <- filter(unemployed, n>200 &
                          ((Election_Type == "General" & 
                              CandidateSurname %in% c("Trump","Clinton")) | 
                             (Election_Type == "Primary")))


ggplot(alluvial_plot,
       aes(y = Total, axis3 = Election_Type, axis1 = Party, 
           axis2 = CandidateSurname)) +
  geom_alluvium(aes(fill=Party), width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label_repel(stat = "stratum", label.strata = TRUE, size=3)+
  scale_x_discrete(limits = c("Party", "Candidate", "Type"), 
                   expand = c(.05, .05)) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  ggtitle("Campaign contributions from Unemployed Contributors")


```

In comparison the the retired, unemployed donors contributed a lot less - just 
over $30 million - but they gave almost all of it to the Democrats. They had an 
extreme preference for Sanders - which is why almost all of their contributions 
were made in the primaries.



# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The two alluvial diagrams show striking differences in the patterns of campaign 
donations by two large an important groups of donors: the retired and the 
unemployed. While the support from retired donors reflects general trends seen 
in the analysis, unemployed donors shunned all Republican candidates and, also 
largely Clinton, to give very strong financial support to Sanders.

The treemap comparing the supporters of the Clinton and Trump general election 
campaigns reveals some interesting aspects. While Clinton receives significantly 
more money during the general election, the median donations by occupation to 
Trump are much higher for a lot of occupations. Also, while some groups are 
shared, others have an aversion to Trump - e.g. teachers, academics, the 
unemployed, those working in the arts. Also, groups that prominently donate to 
both candidates, such as senior managers, doctors, lawyers, donate much higher 
amounts to Trump.



### Were there any interesting or surprising interactions between features?
The violon plots are a good way to compare the dynamic of the Democratic and 
Republican primaries races: after the other candidates left, the Democratic 
primaries became a "two-horse race" between Sanders and Clinton quite early on ; 
by comparisons, the 17 Republican candidates dropped out more gradually, leading
to a more crowded field for much of the context.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
#divide [1,2700] into 8 segments (that would be equal-sized on a log axis)
intervals <- c(0,1,2,3,4,5,6,7,8) * log10(2700)/8 
#divide into 8 bins (log) according to the contribution interval
campaign2016$cuts <- cut(campaign2016$Contribution, 
                         breaks = 10^intervals,
                         include.lowest = TRUE)
Contribution_8bins <- subset(campaign2016,Contribution>=1) %>% #ignore <1
  group_by(cuts) %>%  #create df with 8 bins (log-scale)
  summarise(count = n(), #the count of Contributions in bin
             upper_bound = max(Contribution), #the upper bound of the bin
             sum = sum(Contribution),  #total contributions in bin
             median = median(Contribution)) %>%
    arrange(upper_bound)

p1 <- ggplot(data = Contribution_8bins, 
             aes(x = upper_bound, y = count),
             ylab = "count") +
  geom_col() +
  scale_x_log10(breaks = c(1, 10, 100, 1000, 2700))+
  ylab("count")+
  xlab("Contribution (amount up to, $)")+
  #label outliers/above$2700, adapted from https://stackoverflow.com/a/53448372
  geom_label(aes(label = ifelse(upper_bound > 2700, "above $2700", NA)),
             hjust = 1, vjust = 0)+ 
  ggtitle("Number of contributions per segment (histogram)")

p2 <- ggplot(data = Contribution_8bins, aes(x = upper_bound, y = sum/1000000))+
  geom_col()+
  scale_x_log10(breaks = c(1, 10, 100, 1000, 2700))+
  ylab("total (million $)")+
  xlab("Contribution (amount up to, $)")+
  geom_label(aes(label = ifelse(upper_bound > 2700, "above $2700", NA)),
             hjust = 1, vjust = 0)+
  ggtitle("Income from contributions per segment (million $)")
grid.arrange(p1, p2, ncol = 1)

```

### Description One
The top plot is a histogram representing donations in the range between \$1 and 
\$2,700 into 8 bins (plus an extra bin for donations above \$2,700), while the 
bottom plot visualises the corresponding aggregate sums raised. The disparity 
between the amount raised in the lower bins and the higher bins is stark. The 
first 4 bins raise around $100 million, but a much larger source of campaign 
income -over \$300million- is represented by around 125,000 contributions in the
8th segment of \$1005 to \$2,700. 

This disparity is linked to the log-normal distribution of contributions. It 
expresses an important truth about campaign finance: size of donations matters. 
The amount of political influence that is obtained through campaign finance 
falls dispropotionately on the relatively small number of people who can afford
to donate large amounts.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x = CandidateSurname, y = Date),
       data = subset(campaign2016, 
                     CandidateSurname %in% Rep_final5 &
                       Election_Type == 'Primary')) +
         geom_violin() +
      scale_y_date(date_breaks = "4 month")+
   #primaries start
  geom_hline(aes(yintercept = as.numeric(as.Date('16-02-01'))), linetype=4) +
  #Cruz, Kasich withdraw
  geom_hline(aes(yintercept = as.numeric(as.Date('16-05-03'))), linetype=3) +
   #primaries end
  geom_hline(aes(yintercept = as.numeric(as.Date('16-06-07'))), linetype=4) +
  coord_flip() # time on horizontal axis

```

### Description Two

This plot shows the race between top candidates in the Republican primaries as 
it unfolds. The violin plots show the density of contributions over time 
(normalised to be the same for all candidates). Vertical lines show the 
beginning and the end of the primaries, as well as the point from which Trump 
becomes the sole remaining candidate, after Cruz and Kasich suspended their 
campaigns following the Indiana primaries on 3.May.

Rubio started to receive campaign donations since March 2014, receives the bulk 
of finance around the early part of the primaries - and exits after a month, on 
15. March. Carson - who received a higher proportion of funding in the 6 months 
in the  run up to the primaries start - leaves the race on 4. March. 

By contrast Trump - who partially self-funded his campaign to the tune of $66 
million - received the vast majority of primary contributions after the end 
of the primaries.


### Plot Three
```{r echo=FALSE, Plot_Three}
unemployed <- subset(campaign2016,Occupation == "Unemployed") %>%
    group_by(Election_Type,Party,CandidateSurname) %>%
    summarise(n = n(),
            Total = sum(Contribution)/10^6,
            Mean = mean(Contribution),
            Median = median(Contribution)) %>%
  arrange(Party) %>%
  ungroup()

alluvial_plot <- filter(unemployed, n>200 &
                          ((Election_Type == "General" & 
                              CandidateSurname %in% c("Trump","Clinton")) | 
                             (Election_Type == "Primary")))


ggplot(alluvial_plot,
       aes(y = Total, axis3 = Election_Type, axis1 = Party, 
           axis2 = CandidateSurname)) +
  geom_alluvium(aes(fill=Party), width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label_repel(stat = "stratum", label.strata = TRUE, size=3)+
  scale_x_discrete(limits = c("Party", "Candidate", "Type"), 
                   expand = c(.05, .05)) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  ylab("total (million $)") +
  ggtitle("Campaign contributions from Unemployed Contributors")

```

### Description Three
This alluvial diagram shows what is perhaps the strongest relationship revealed 
in this analysis: donors who are unemployed show strong support for Sanders 
specifically. Of the $30 million they donate, almost nothing goes to Republican
candidates, and perhaps 1/10 to Clinton. Similarly the majority of contributions
are in the primaries: after Sanders leaves the race, they lose almost all 
interest in supporting the Democrats financially.
------

# Reflection

Really interesting to learn some new types of plots like treemaps and alluvial 
diagrams. Also interesting to analyse in more depth the distribution of 
contribution amounts and the reason behind 'negative contributions'.

Cleaning the donor occupation data was more work than I had anticipated but
once it was done it made the analysis pretty interesting. Frustratingly, a lot 
of the data remains un-analysed.

Perhaps it may seem questionable to choose the whole dataset of the 2016 
election campaign, rather than only data for a single state. My original plan 
had been to also use the ZIP codes to map donors geographically, and combine 
this with information such as deprivation, unemployment, ethnic diversity, 
rural vs urban character. Trump (and to some extent Sanders) have been portrayed
as 'disruptors' of the normal rules of US elections - it would have been very 
intersting to look for evidence of them mobilising particular demographics in
particular areas, perhaps attempt to build a predictive model for donors in
which neighbourhoods unexpectedly support a particular candidate. Unfortunately
the amount of work required turned out to be beyond the scope of this project.

### References 

the following is a list of Web sites used in creating my submission 

https://transition.fec.gov/pages/brochures/citizens.shtml
http://fec.gov/disclosurep/pnational.do&sa=D&ust=1548605508118000
https://www.opensecrets.org/pres16/candidate?id=N00023864
https://en.wikipedia.org/wiki/Hillary_Victory_Fund)
https://en.wikipedia.org/wiki/2016_Democratic_Party_presidential_primaries)
https://en.wikipedia.org/wiki/2016_Republican_Party_presidential_primaries)
https://en.wikipedia.org/wiki/Swing_state
https://stackoverflow.com/a/47210732
https://github.com/datasciencedojo/DataMiningFEC/blob/master/6%20Bucketing%20Occupation%20Groups.R
https://gist.github.com/jennybc/6f3fa527b915b920fdd5
https://stackoverflow.com/a/31925893
https://stackoverflow.com/a/53448372
https://stackoverflow.com/a/17374663
https://stackoverflow.com/a/51147233
https://stackoverflow.com/a/42237896
https://stackoverflow.com/a/45820586
https://stackoverflow.com/a/22181949
https://ggplot2.tidyverse.org/reference/geom_bar.html
https://stackoverflow.com/a/6901597
